---
title: "ESM 244 Lab 1"
author: "Tiffany Hsu"
date: "1/6/2022"
output: html_document
---

```{r setup, include = TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
library(sf)
library(tmap)

# to update packages use `update.packages(ask = FALSE)`
```

### Reading in data

```{r}
sf_trees <- read_csv(here("data", "sf_trees", "sf_trees.csv"), show_col_types = FALSE) # wont show the column types
```

### Part 1: wrangling and ggplot review

**Example 1:** Find counts of observations by `legal_status` & wrangle a bit.

```{r}
### method 1: `group_by() %>% summarize()
sf_trees %>% 
  group_by(legal_status) %>% 
  summarize(tree_count = n())

### method 2: different way plus a few new functions
top_5_status <- sf_trees %>% 
  count(legal_status) %>% 
  drop_na(legal_status) %>% 
  rename(tree_count = n) %>% 
  relocate(tree_count, 1) %>% # reorder columns
  slice_max(tree_count, n =5) %>% # takes top 5 highest values
  arrange(desc(tree_count)) # highest to lowest value sort
```

Make a graph of the top 5 from above
```{r}
ggplot(data = top_5_status, aes(x = fct_reorder(legal_status, tree_count), #fct_reorder orders from smallest to largest # of trees
                                y = tree_count)) +
  geom_col(fill = 'darkgreen') +
  labs(x = 'Legal status',
       y = 'Tree count') +
   coord_flip() + #will flip the axis labels so they fit the entire word
  theme_minimal()
  
```

**Example 2:** Only going to keep observations where legal status is "Permitted Site" and caretaker is "MTA", and store as `permitted_data_df`

```{r}
# sf_trees$legal_status %>% unique() # checks for unique values
permitted_data_df <- sf_trees %>% 
  filter(legal_status == "Permitted Site", 
         caretaker == "MTA")
```

**Example 3:** Only keep Blackwood Acacia trees, and then only keep columns `legal_status`, `date`, `latitude`. and store as `blackwood_acacia_df`
```{r}
blackwood_acacia_df <- sf_trees %>% 
  filter(str_detect(species, 'Blackwood Acacia')) %>% 
  select(legal_status, date, lat = latitude, lon = longitude)

# Make a little graph of locations
ggplot(data = blackwood_acacia_df, aes(x = lon, y = lat)) +
  geom_point(color = "darkgreen") 
```

**Example 4:** use `tidyr::separate()` to separate words in a column into two separate columns

```{r}
sf_trees_map <- sf_trees %>% 
  separate(species, into = c('spp_scientific', 'spp_common'), sep = '::')
```

**Example 5:** use `tidyr::unite()`

```{r}
ex_5 <- sf_trees %>% 
  unite('id_status', tree_id, legal_status, sep = ' ADDING THIS ')
```

### Part 2: make some maps

**Step 1:** convert the lat/lon to spatial points, `st_as_sf()`

```{r}
blackwood_acacia_sf <- blackwood_acacia_df %>% 
  drop_na(lat, lon) %>% 
  st_as_sf(coords = c('lon', 'lat'))

# we need to tell R what the coordinate reference system is
st_crs(blackwood_acacia_sf) <- 4326 #WGS84

ggplot(data = blackwood_acacia_sf) +
  geom_sf(color = 'darkgreen') +
  theme_minimal()
```

Read in the SF shapefile and add to map
```{r}
sf_map <- read_sf(here("data", "sf_map", "tl_2017_06075_roads.shp"))

sf_map_transform <- st_transform(sf_map, 4326)

ggplot(data = sf_map_transform) +
  geom_sf()
```

Combine the maps!

```{r}
ggplot() +
  geom_sf(data= sf_map,
          size = 0.1,
          color = "darkgrey") + #will be on the bottom layer 
  geom_sf(data= blackwood_acacia_sf,
          size = 0.5,
          color = "darkgreen") +
  theme_void() +
  labs(title = "Blackwood Acacias in SF")
```

### Now an interactive one! 

```{r}
tmap_mode("view")

tm_shape(blackwood_acacia_sf) + 
  tm_dots()
```




